---
title: "Assignment 4"
author: "Ashani Jayasekera"
date: "2024-10-23"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(DBI)
library(dplyr)
```

## Assignment 4

The link to the GitHub repository is: https://github.com/ashanijay/SURV727. The assignment is Assignment4.rmd.

```{r cars1}
project <- "surv727-hw4-ajayase1"

con <- dbConnect(
  bigrquery::bigquery(),
  project = "bigquery-public-data",
  dataset = "chicago_crime",
  billing = project
  )
con
dbListTables(con)


```

## Including Plots

You can also embed plots, for example:

```{sql connection = con}
SELECT count(primary_type), count(*)
FROM crime
WHERE year = 2016
LIMIT 10
```
The number of arrests grouped by primary_type in 2016 is in the table below. 
```{sql connection= con, message = FALSE}
SELECT primary_type,
      COUNTIF(year = 2016) AS arrests,
FROM crime
WHERE year = 2016
GROUP BY primary_type
ORDER BY COUNTIF(year = 2016) DESC
LIMIT 10

```

The date is now used for grouping instead of primary_type.
```{sql connection= con, message = FALSE}
SELECT date,
      COUNTIF(year = 2016) AS arrests,
FROM crime
WHERE year = 2016
GROUP BY date
ORDER BY COUNTIF(year = 2016) DESC
LIMIT 10

```
The number of arrests grouped by hour of the day in 2016 is below. 
```{sql connection= con, message = FALSE}
SELECT EXTRACT(HOUR FROM date),
      COUNTIF(year = 2016) AS arrests,
FROM crime
WHERE year = 2016
GROUP BY EXTRACT(HOUR FROM date)
ORDER BY COUNTIF(year = 2016) DESC
LIMIT 10

```

Now the arrests focused on homicide are below ordered by year in descending order. 
```{sql connection= con, message = FALSE}
SELECT year,
      COUNTIF(primary_type = "HOMICIDE") AS arrest,
FROM crime
WHERE primary_type = "HOMICIDE"
GROUP BY year
ORDER BY COUNTIF(primary_type = "HOMICIDE") DESC
LIMIT 10


```
The districts with the highest numbers of arrests in 2015 and 2016 are below. They are grouped by year and district. 
```{sql connection= con, message = FALSE}
SELECT year, district,
      COUNTIF(year = 2015) AS arrests_2015,
      COUNTIF(year = 2016) AS arrests_2016
FROM crime
WHERE year = 2015 OR year = 2016
GROUP BY district, year
ORDER BY COUNTIF(year = 2015) DESC, COUNTIF(year = 2016) DESC


```
Now the code is in R via the DBI package. 
The number of arrests grouped by primary_type of district 11 in year 2016 is below. 
```{r cars3}
query <- "
  SELECT primary_type, COUNT(*) AS arrests
  FROM crime
  WHERE year = 2016 AND district = 11
  GROUP BY primary_type
  ORDER BY arrests DESC
  LIMIT 10
"

dbGetQuery(con, query)

```
The same query is now written using the dbplyr package. 

```{r cars4}
library(dplyr)
crime_tbl <- tbl(con, "crime")

arrest_data <- crime_tbl %>%
  filter(year == 2016, district == 11) %>%
  group_by(primary_type) %>%
  summarise(arrests = n()) %>%
  arrange(desc(arrests)) %>%
  head(10) %>%
  collect() 

print(arrest_data)
```

Count the number of arrests grouped by primary_type and year, still only for district 11.
Arrange the result by year. The results are below. 

```{r cars5}

arrest_data2 <- crime_tbl %>%
  filter(district == 11) %>%
  group_by(primary_type, year) %>%
  summarise(arrests = n()) %>%
  arrange(desc(year)) %>%
  head(10) %>%
  collect() 

print(arrest_data2)
```
The connection is now closed. 
```{r cars5}
dbDisconnect(con)
```


