---
title: "Assignment5"
author: "Ashani Jayasekera"
date: "2024-11-13"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(censusapi)
library(tidyverse)
library(magrittr)
library(factoextra)
```
## Assignment 5

The link to the GitHub repository is: https://github.com/ashanijay/SURV727. The assignment is Assignment5.rmd.

```{r}
cs_key <- "65c0979cfd3f4f5585b3572df63418505d19b6e7"
```

## Exploring ACS Data

```{r}
acs_il_c <- getCensus(name = "acs/acs5",
 vintage = 2016,
 vars = c("NAME", "B01003_001E", "B19013_001E", "B19301_001E"),
 region = "county:*",
 regionin = "state:17",
 key = cs_key) %>%
 rename(pop = B01003_001E,
 hh_income = B19013_001E,
 income = B19301_001E)
 
head(acs_il_c)

```

Pull the map data for Illinois into a data frame. 

```{r}
il_map <- map_data("county", region = "illinois")
head(il_map)
```
Join the ACS data with the map data. Call the joined data acs_map. 
```{r}
acs_il_c$subregion <-tolower(sapply(strsplit(acs_il_c$NAME, " County,"), function(x) x[1]))
acs_map <- left_join(acs_il_c, il_map, by = "subregion")
```

Plot a mpa of Illinois with counties colored per capita income. 
```{r}
ggplot(acs_map) +
 geom_polygon(aes(x = long, y = lat, group = group, fill = income))
```

## Hierarchial Clustering
Clean the data and then create distance matrix of the cleaned data. We want to find clusters of ounties that are similar in their population, average household income and per capita income. 

Plot the dendrogram to find a reasonable number of clusters. Draw boxes around the clusters of your cluster solution

```{r}
acs_map_clean <- acs_map %>%
  select(pop, hh_income, income) %>%
  mutate_all(scale)
acs_d <- dist(acs_map_clean)
acs_ward <- hclust(acs_d, method = "ward.D2")
plot(acs_ward$height, main = "Agglomeration Coefficients", xlab = "Number of Merges", ylab = "Height")

```

Plot the dendrogram to find a reasonable number of clusters. Draw boxes around the clusters of your cluster solution.
```{r}
plot(acs_ward, main = "Ward", xlab = "", sub = "")

rect.hclust(acs_ward, 
            k = 6, 
            border = "red")
```

```{r}

##need to find the way to get the optimal number of groups
cutree(acs_ward, 6)
```


Visualize the county clusters on a map. Foe this task, create a new acs_map object that now includes cluster membership as a new column. This column should be called cluster. 
```{r}

acs_map$cluster <- as.factor(cutree(acs_ward, 6))

ggplot(acs_map) +
  geom_polygon(aes(x = long, y = lat, group = group, fill = cluster))
```
## Census Tracts
```{r}
acs_il_t <- getCensus(
  name = "acs/acs5",
  vintage = 2016,
  vars = c("NAME", "B01003_001E", "B19013_001E", "B19301_001E"),
  region = "tract:*",
  regionin = "state:17",  # Illinois (state FIPS = 17)
  key = cs_key
) %>%
  # Replace -666666666 with NA in all columns
  mutate(across(everything(), ~ ifelse(. == -666666666, NA, .))) %>%
  # Rename columns
  rename(
    pop = B01003_001E,
    hh_income = B19013_001E,
    income = B19301_001E
  )
acs_il_t <- acs_il_t[complete.cases(acs_il_t), ]
head(acs_il_t)
```

## k-Means
Clean the data for clustering census tracts pased on population, average hourshold income, and per capita income. 

Since we want to use K Means in this section, we start by determining the optimal number of K that results in Clusters with low within but high between variation. Plot within cluster sums of squares for a range of K (e.g. up to 20).

Run kmeans() for the optimal number of clusters based on the plot above.

```{r}
acs_il_clean <- acs_il_t %>%
  select(pop, hh_income, income) %>%
  mutate_all(scale)
acs_il_clean <- acs_il_clean[complete.cases(acs_il_clean), ]
  

fviz_nbclust(acs_il_clean, #data set we want to use
             kmeans, #cluster method
             method = "wss", #method used for estimating the optimal number of clusters
             k.max = 20)
```
Based on the plots, the optimal number of clusters is either 2, 3, or 6. It is not clear. But the best I think is 6. The cluster means for population, household income, adn per capita income grouped by cluster is below. 
```{r}
km_1 <- kmeans(acs_il_clean, 6, nstart = 20)
km_1
```
In addition, display the most frequent county that can be observed within each cluster. 
```{r}
acs_il_t$cluster <- km_1$cluster
result <- acs_il_t %>%
  group_by(cluster, county) %>%  
  count() %>%  
  group_by(cluster) %>%
filter(n == max(n)) %>%  
  ungroup()  

 print(result)
```
To automate K Means clustering, program a function based on kmeans() that takes K as an argument.
```{r}
kmeans_cluster <- function(K, dataset.clean, dataset.full){
    cluster_id <- kmeans(dataset.clean, K, nstart = 20)$cluster
  name <- paste0("clusterID_", K)
  dataset.full <- cbind(dataset.full, cluster_id)
  names(dataset.full)[dim(dataset.full)[2]] <- name
  return(dataset.full)
  }

for(k in 2:10){
  acs_il_t <- kmeans_cluster(K = k, dataset.clean = acs_il_clean, dataset.full = acs_il_t)
}
head(acs_il_t)
```
